\bf No More, No Less Than Sum of its Parts: The Algebra of Graphics, Statistics, and Interaction

### 2024-10-19 paper.tex
Wrote some more

The essence of data visualization is comparison. When we visualize, we
split our data into parts, summarize each part by a set of one or more
statistics, and then encode these summaries into visual channels such as
position, size, or color that we then use to draw geometric objects
\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
example, when drawing a typical barplot, we split our data based on the
levels of some categorical variable, count or sum up the values of some
other variable, and finally represent these summaries by drawing bars of
the corresponding heights. In this way, each bar becomes a visual
representation of its own ``small data set'', a subset of the original
data.

To ground this idea in a more concrete example, one that will be used
throughout the rest of this paper, imagine going through your bookshelf,
sorting the books by genre, and stacking them on top of each other. You
would construct a physical version of the barplot, forming one ``bar''
(stack) per genre: one bar for sci-fi novels, one bar for romance, one
bar for detective fiction, and so on. Even from this very simple
example, interesting mathematical properties jump out: for example, a
single book cannot go into two stacks (the subsets of the data are
\emph{disjoint}) and if all of the stacks are combined back together, we
should end up with the bookshelf we started with \citep[the subsets are
\emph{surjective} and cover the entirety of the data set,
see][]{ziemkiewicz2009}.

Now, just as we can learn about the contents of our library by
organizing books in the way we just described, we can learn about our
data by splitting it into parts. A natural question that arises is: can
we learn more by subdividing our data into even smaller parts and
comparing these? For instance, in our book example, we could split each
stack of books into two smaller stacks based on the gender of the
author. Each smaller stack will then represent the unique combination
(Cartesian product) of the two categorical variables (genre-gender).
Moreover, there arises a hierarchical relationship between the stacks:
by combining two genre-gender stacks (child components) together, we can
recover the original genre stack (parent component). Similar situation
often arises in interactive graphics, where we use techniques such as
linked selection to highlight specific subsets of our data. Either way,
by forming these finer partitions, we can represent our data with
greater granularity and discover trends that may otherwise hide in the
aggregate.

However, once we have subdivided each category into its child
components, we need a way of representing this information that visually
preserves the hierarchical nature of the relationship. In data
visualization, there are two popular methods for doing this, known as
``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
methods have existed at least since the time of William Playfair
\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
briefly illustrate them on the example of the classical barplot. In a
stacked barplot, bar segments representing the product of the two factor
variables are plotted vertically on top of each other (``stacked''; the
operation can be also thought of as ``highlighting'' parts of the bar).
In contrast, in a dodged barplot, bars are plotted side-by-side, as
clusters of bars. Finally, there is also a third popular method, known
as layering, where we plot bar segments in separate graphical layers and
use partial transparency to mitigate overplotting (this technique may
have been a bit challenging for William Playfair but has been made
simple with computers).

@@

Much has been written about the relative merits of stacking, dodging,
and layering. For example, layering is only useful with few categories,
as blending many colors can make it difficult to tell the categories
apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
\citet{cleveland1984} showed that people tend to be less accurate when
reading information from stacked bar charts as opposed to dodged bar
charts. Specifically, since the lower y-axis coordinate of a stacked
segment is pushed up by the cumulative height of the segments below it,
it becomes difficult to accurately compare segments' length, both within
and across bars \citep{cleveland1984}. Subsequent research has
independently validated and expanded upon these findings \citep[see
e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
statistical legibility of stacking has led some authors to advise
against its use altogether \citep{kosara2016, wilke2019}, or at least
urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.

However, despite the mixed reception of stacked plots in static
visualization, they remain a popular choice in interactive data
visualization. Among popular interactive data visualization libraries,
stacked plots such as bar charts and histograms make frequent appearance
\citep[see
e.g.][]{cook2023, murray2017, vegalite2022, vanderplas2018, sievert2020, swayne2003, xie2014, theus2002, urbanek2003, urbanek2011, wills2008, ware2019}.
What is the reason for this? Has the interactive data visualization
community simply failed to catch up with the current best practices? Or
is there perhaps more going on?

We argue that not only is stacking a useful technique, but that it can
in fact offer valuable insights about the mathematical structure
underlying our visualizations. To make the strongest case possible, we
bring forward evidence from several levels of the visualization process.
We start by discussing the properties of visual perception and how they
apply to both static and interactive data visualizations. Next, we
explore the process of computing statistical summaries that form the
basis of our plots. Finally, we dive into the algebraic properties of
the functions underlying these statistical summaries. Throughout the
text, we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and
efficient interactive graphics, we simply cannot treat graphics,
statistics, and interaction as independent components. Instead we need
to consider them holistically, within the context of their respective
visual and mathematical properties. Understanding this deep web of
interdependence may in turn empower us to innovate and build new kinds
of robust interactive visualizations.

@@

Whether we intend them to or not, the Gestalt principles how we perceive
data visualizations. As such, it is a good idea to try to use them to
our advantage. Sometimes, little effort is needed. Such is the case, for
example, when identifying clusters in a scatterplot \citep{rosli2015}:
the principles of proximity and similarity ensure that clouds of
similarly shaped and colored points will be perceived as discrete groups
by the observer. In other cases, however, our design choices can have a
significant impact. For instance, another way to have objects to be
perceived as part of a group is to draw a closed shape around them
\citep{cairo2012}. Alternatively, if we want to highlight a particular
part of a data set, we can leverage the figure and ground principle and
use attributes such as color and saturation to make it stand out
\citep[for specific examples, see][34]{franconeri2021, wilke2019}.

Finally, ignoring or violating the Gestalt principles can make our
graphics less legible or even misleading. \citet{tufte2001} had
documented many such violations, such as deceptive 3D effects,
inadvertent optical illusions caused by dense stripe patterns, and the
overuse of superfluous graphical elements he termed ``chartjunk''.
Similarly, \citet{cairo2019} highlights many cases of graphics that
violate the Gestalt principles. Finally, there is also the issue of
well-designed graphics that communicate no statistical information at
all, or even information that is blatantly false
\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
Either way, whether intentional or not, poor design can have a
significant impact on the legibility and effectiveness of statistical
graphics.

@@

Linked selection, also known as linked brushing or linked highlighting,
has been consistently ranked as one of the most useful interactive
features in data visualization \citep[see
e.g.][]{buja1996, heer2012, wilhelm2003, ware2019, ward2015}. By
clicking or clicking-and-dragging over objects in one plot, users can
highlight parts of objects corresponding to the selected cases across
all other ``linked'' plots. This allows the user to rapidly
``drill-down'' \citep{dix1998, theus2002} and explore trends across
different dynamically-generated subsets of the data \citep[similar to
interactive filtering, see e.g.][]{heer2012}. The ability to rapidly
materialize different views of the data makes linked selection a
particularly versatile tool for data exploration.

Yet, despite its versatility, linked selection also imposes some
fundamental constraints on the visualization. Specifically, all objects
we visualize need to be able to:

@@

Dodged barplots lack the degree of visual consistency found in stacked
barplots. In a dodged barplot, the clusters of segments placed
side-by-side lack a shared border (common region), leading to several
design challenges. First, when selection occurs, the jagged contour of
the dodged barplot can change dramatically as more cases are added or
removed. This creates distracting visual clutter that makes it difficult
to track changes over time. Second, selection may alter the height of
the tallest bar segment in the plot. Since the height of the bar
segments is not bound by the height of the tallest bar (as is the case
in stacked barplot), this presents us with a dilemma. Either either we
make the upper y-axis limit dynamic, meaning that the bar segments will
always fit the plotting region, but then the context that the upper
y-axis limit provides is lost with each time selection happens.
Alternatively, we keep the upper y-axis limit static, but then we risk
the bar segments growing outside of the bounds of the plotting region.

>>>>

The essence of data visualization is comparison. When we visualize, we
split our data into parts, summarize each part one or more summary
statistics, encode the summaries into visual attributes such as
position, size, or color, and finally use these attributes to draw
geometric objects
\citetext{\citealp{bertin1983}; \citealp{wilkinson2012}; \citealp[for a
recent review, see][]{franconeri2021}; \citealp{wilke2019}}. For
example, when drawing a typical barplot, we split our data into parts
based on the levels of some categorical variable, count the number of
cases in each part, and finally represent the counts by drawing bars of
the corresponding heights. In this way, each bar becomes a visual
representation of its own ``small data set'', a subset of the original
data.

To ground this idea in a more concrete example that will be used
throughout this paper, imagine going through your bookshelf, sorting the
books by genre, and stacking them on top of each other. You would
construct a physical version of the barplot, forming one ``bar'' (stack)
per genre: one bar for sci-fi novels, one bar for romance, one bar for
detective fiction, and so on. Even from this very simple example,
interesting mathematical properties jump out: for example, a single book
cannot go into two stacks (the subsets of the data are \emph{disjoint})
and if all of the stacks are combined back together, we should end up
with the bookshelf we started with \citep[the subsets are
\emph{surjective} and cover the entirety of the data set,
see][]{ziemkiewicz2009}.

Now, just as we can learn about the contents of our library by
organizing books in the way we just described, we can learn about our
data by splitting it into parts. A natural question then arises: can we
learn more by subdividing our data into even smaller parts and comparing
these? For instance, in our book example, we can split each stack of
books into two smaller stacks based on the gender of the author. Each
smaller stack will then represent the unique combination (Cartesian
product) of the two categorical variables (genre and gender). Moreover,
there arises a hierarchical relationship between the stacks: by
combining two genre-gender stacks (child components) together, we can
recover the original genre stack (parent component). Similar situation
often arises in interactive graphics, where we use techniques such as
linked selection to highlight specific subsets of our data. Either way,
by forming these finer partitions, we can represent our data with
greater granularity and discover trends that may otherwise hide in the
aggregate.

However, once we have subdivided each category into its child
components, we need to represent the information in a way that visually
preserves the hierarchical nature of this relationship. In data
visualization, there are two popular methods for doing this, known as
``stacking'' and ``dodging'' (see Figure \ref{fig:playfair}). These
methods have existed at least since the time of William Playfair
\citetext{\citeyear{playfair1801}; \citeyear{playfair1822}}. Let's
briefly illustrate them on the example of the classical barplot. In a
stacked barplot, bar segments representing the product of the two factor
variables are plotted vertically on top of each other (``stacked''; the
operation can be also thought of as ``highlighting'' parts of the bar).
In contrast, in a dodged barplot, bars are plotted side-by-side, as
clusters of bars. Finally, there is also a third popular method, known
as layering, where we plot bar segments in separate graphical layers and
use partial transparency to mitigate overplotting (this technique may
have been a bit challenging for William Playfair but has been made
simple with computers).

@@

Much has been written about the relative merits of stacking, dodging,
and layering. For example, layering is only useful with few categories,
as blending many colors can make it difficult to tell the categories
apart \citep{franconeri2021, wilke2019}. Further, in a landmark study,
\citet{cleveland1984} showed that people tend to be less accurate when
reading information from stacked bar charts as opposed to dodged bar
charts. Specifically, since the lower y-axis coordinate of a stacked
segment is pushed up by the cumulative height of the segments below it,
it becomes difficult to accurately compare segments' length, both within
and across bars \citep{cleveland1984}. Subsequent research has
independently validated these findings and expanded upon them \citep[see
e.g.][]{heer2010, thudt2016, quadri2021}. Ultimately, the suboptimal
statistical legibility of stacking has led some authors to advise
against its use altogether \citep{kosara2016, wilke2019}, or at least
urge caution \citep[see e.g.][]{byron2008, cairo2014, franconeri2021}.

However, despite the mixed reception towards stacking in static
visualization, it remains a popular choice in interactive data
visualization. Among popular interactive data visualization libraries,
stacked plots such as bar charts and histograms make frequent appearance
\citep[see
e.g.][]{cook2023, murray2017, vegalite2022, vanderplas2018, sievert2020, swayne2003, xie2014, theus2002, urbanek2003, urbanek2011, wills2008, ware2019}.
What is the reason for this? Has the interactive data visualization
community simply failed to catch up with the current best practices? Or
is there perhaps more going on?

We argue that not only is stacking a useful technique, but that it can
in fact offer valuable insights about the mathematical structure
underlying our visualizations. To make the strongest case possible, we
bring forward evidence from several levels of the visualization process.
We start by discussing the properties of visual perception and how they
apply to both static and interactive visualizations. Next, we explore
the process of computing statistical summaries that form the basis of
our plots. Finally, we dive into the algebraic properties of the
functions underlying these statistical summaries. Throughout the text,
we use the example of the humble barplot to ground our discussion.

Our goal is to convince you that, in order to produce coherent and
efficient interactive graphics, we simply cannot treat graphics,
statistics, and interaction as independent components. Instead we need
to consider them holistically, within the context of their respective
visual and mathematical properties. Understanding this deep web of
interdependence may in turn empower us to innovate and build new kinds
of robust and expressive interactive visualizations.

@@

Whether we intend them to or not, the Gestalt principles affect how our
data visualizations are perceived. As such, it is a good idea to try to
use them to our advantage. Sometimes, little effort is needed. Such is
the case, for example, when identifying clusters in a scatterplot
\citep{rosli2015}: the principles of proximity and similarity ensure
that clouds of similarly shaped and colored points will be perceived as
discrete groups by the observer. In other cases, however, our design
choices can have a significant impact. For instance, another way to have
objects to be perceived as part of a group is to draw a closed shape
around them \citep{cairo2012}. Alternatively, if we want to highlight a
particular part of a data set, we can leverage the figure and ground
principle and use attributes such as color and saturation to make it
stand out \citep[for specific examples,
see][34]{franconeri2021, wilke2019}.

Finally, ignoring or violating the Gestalt principles can make our
graphics less legible or even misleading. \citet{tufte2001} had
documented many such violations, such as deceptive 3D effects,
inadvertent optical illusions caused by dense stripe patterns, and the
overuse of superfluous graphical elements he termed ``chartjunk''.
Similarly, \citet{cairo2019} highlights many cases of graphics that
violate the Gestalt principles. Finally, there is also the issue of
well-designed graphics that communicate no statistical information at
all, or even information that is blatantly false
\citep{cairo2014, franconeri2021, cairo2019, gelman2013, tufte2001}.
Whether intentional or not, poor design can have a significant impact on
the legibility and effectiveness of statistical graphics.

@@

Linked selection, also known as linked brushing or linked highlighting,
has been consistently ranked as one of the most useful interactive
features in data visualization \citep[see
e.g.][]{buja1996, heer2012, wilhelm2003, ware2019, ward2015}. By
clicking or clicking-and-dragging over objects in one plot, users can
highlight parts of objects corresponding to the selected cases across
all other ``linked'' plots. This allows the user to rapidly
``drill-down'' \citep{dix1998, theus2002} and explore trends across
different dynamically-generated subsets of the data \citep[similar to
interactive filtering, see e.g.][]{heer2012}. The ability to quickly
materialize different views of the data makes linked selection a
versatile tool for data exploration.

Yet, despite its usefulness, linked selection also imposes some
fundamental constraints on the figure. Specifically, all objects we
visualize need to be able to:

@@

Dodged barplots lack this degree of visual consistency. In a dodged
barplot, bar segments placed side-by-side in clusters, and, since each
segment can be of different length, these clusters lack a shared border
(common region). This leads to several design challenges. First, when
selection occurs, the jagged contour of the dodged barplot can change
dramatically as more cases are added or removed. This creates
distracting visual clutter that makes it difficult to track changes over
time. Second, since the heights of the bar segments are not bound by any
value present in the plot (unlike the heights of the segments in the
stacked barplot, which are bound by the height of the tallest bar), we
need to decide how the handle the upper y-axis limit in response to
selection. One option is to make the upper y-axis limit dynamic. If we
do this, we ensure that the bar segments always fit the plotting region,
however, we also lose the context that the upper y-axis limit provides
each time selection occurs. Alternatively, we can assign a static value
to the upper y-axis limit, however, then we risk the bar segments
growing outside of the bounds of the plotting region.
